The Astropy Project: \\ Sustaining and Growing a Community-oriented Open-source Project and the Latest Major Release (v5.0) of the Core Package

### 2022-02-23 src/ms.tex
Few more comments
Addressing comments, thanks @cosimoNigro, @matteobachetti, @sibirrer

% Links to documentation: subpackages
\newcommand{\astropysubpkg}[1]{\href{http://docs.astropy.org/en/stable/#1/index.html}{\texttt{astropy.#1}}\xspace}
\newcommand{\astropyiosubpkg}[1]{\href{http://docs.astropy.org/en/stable/io/#1/index.html}{\texttt{astropy.io.#1}}\xspace}
\newcommand{\astropywcsaxes}{\href{http://docs.astropy.org/en/stable/visualization/wcsaxes/index.html}{\texttt{astropy.visualization.wcsaxes}}\xspace}
\newcommand{\astropycosmology}{\astropysubpkg{cosmology}}
\newcommand{\astropycosmologyunits}{\href{https://docs.astropy.org/en/stable/cosmology/units.html}{\texttt{astropy.cosmology.units}}}
\newcommand{\astropyunits}{\astropysubpkg{units}}
\newcommand{\astropycoordinates}{\astropysubpkg{coordinates}}
\newcommand{\astropyconstants}{\astropysubpkg{constants}}
\newcommand{\astropytime}{\astropysubpkg{time}}
\newcommand{\astropymodeling}{\astropysubpkg{modeling}}

% Links to documentation: classes
\newcommand{\astropyapi}[2]{\href{https://docs.astropy.org/en/stable/api/astropy.#1.html}{#2}}
\newcommand{\astropyapidoc}[2]{\astropyapi{#1}{\texttt{#2}\xspace}}

\newcommand{\astropySpectralCoord}{\astropyapidoc{coordinates.SpectralCoord}{SpectralCoord}}
\newcommand{\astropySkyCoord}{\astropyapidoc{coordinates.SkyCoord}{SkyCoord}}
\newcommand{\astropyICRS}{\astropyapidoc{coordinates.builtin_frames.ICRS}{ICRS}}
\newcommand{\astropyGalacticLSR}{\astropyapidoc{coordinates.builtin_frames.
GalacticLSR}{GalacticLSR}}
\newcommand{\astropyAltAz}{\astropyapidoc{coordinates.builtin_frames.AltAz}{AltAz}}
\newcommand{\astropyGalactocentric}{\astropyapidoc{coordinates.builtin_frames.Galactocentric}{Galactocentric}}
\newcommand{\astropyCosmology}{\astropyapidoc{cosmology.Cosmology}{Cosmology}}
\newcommand{\astropyFlatLambdaCDM}{\astropyapidoc{cosmology.FlatLambdaCDM}{FlatLambdaCDM}}
\newcommand{\astropyTable}{\astropyapidoc{table.Table}{Table}}
\newcommand{\astropyQTable}{\astropyapidoc{table.QTable}{QTable}}
\newcommand{\astropyTime}{\astropyapidoc{time.Time}{Time}}
\newcommand{\astropyLeapSeconds}{\astropyapidoc{time.LeapSeconds}{LeapSeconds}}
\newcommand{\astropyDistribution}{\astropyapidoc{uncertainty.Distribution}{Distribution}}
\newcommand{\astropyUnit}{\astropyapidoc{units.Unit}{Unit}}
\newcommand{\astropyQuantity}{\astropyapidoc{units.Quantity}{Quantity}}
\newcommand{\astropyMasked}{\astropyapidoc{utils.masked.Masked}{Masked}}
\newcommand{\astropyScienceState}{\astropyapidoc{utils.state.ScienceState}{ScienceState}}

% suggestion: shortcut reference to Paper II
\newcommand{\paperii}{\cite{astropy:2018}}

% Maybe have command link to Zenodo reference instead, or include it?
\newcommand{\astropyAPE}[1]{\href{https://github.com/astropy/astropy-APEs/blob/main/APE#1.rst}{APE #1}\xspace}

@@

The \astropypkg core package is mature and stable, in that many features have
been part of the core package for years with (purposefully) few changes to the
software interface. This maturity and stability allows for the broader
astronomy \python community to rely upon the \astropypkg core package and
build specialized packages within the Astropy ecosystem. Even many
non-astronomical \python libraries have come to rely on \astropypkg.

The relationship between \astropy and community is reciprocal, with the
growing and evolving needs of the community and the maturation of the broader
scientific \python ecosystem now driving the development of the \astropypkg
core package.

One manifestation of this was the realization that some features do not actually
belong in the core package, either because they are of more general use
outside of \astropypkg, or because they are too specialized. An example of the
former is the copy of the IAU Standards Of Fundamental Astronomy (SOFA)
software Collection\footnote{\url{http://www.iausofa.org}} \citep{sofa} that
\astropy carried, as well as the \python wrappers for it.  Since this is more
basic infrastructure, with a release cycle set by SOFA, it made more sense to
create separate packages that \astropypkg can depend on: \package{ERFA}
\citep{erfa} and \package{PyERFA} \citep{pyerfa}. An example of something that
was too specialized was a Virtual Observatory sub-package (now removed).  Of
its content, the Simple Application Message Protocol made sense to keep
(\astropysubpkg{samp}), but other parts belonged more properly in {\tt
Astroquery} (see Sect.~\ref{sec:astroquery}).

Within \astropypkg, development may be roughly split into a few
categories: new features added, intra and inter-package
interoperability, improvements to precision, accuracy, and
reproducibility, and performance enhancements. We discuss these in turn.

\subsubsection*{New and Planned Features} \label{sec:core-features-new}

% \begin{itemize}
%  \item Time Series sub-package (V3.1: New Box Least Squares Periodogram)
%  \item Uncertainties and Distributions
%   \item ERFA module is now it’s own thing; Obliquely tie in that units
%    remains in astropy -> above
%  \item ? Pytest plugin ecosystem -> done elsewhere
%  \item A new SpectralCoord class for representing and transforming spectral quantities
%    Velocity in Coordinates
%  \item WCS API : A shared Python interface for World Coordinate Systems
%        Include APE 14 reference
% \end{itemize}

New features come in various forms, from new submodules providing
wholly new capabilities, to larger additions to existing submodules,
to large rewrites of the interface of modules.  This range is spanned
by the five main new features that entered the astropy core since the
previous summary \citepalias{astropy:2018}.

New features present a particular difficulty for a mature package such
as \astropypkg, because by their nature they have not been tried out
much and it is difficult to know in advance what the best interface
will be, and hard to address all possible use cases in one go, even
within \astropypkg\ itself.  To give a sense of this, we include in
our descriptions how we hope the new features will evolve (realizing
that, in the end, in a user-driven project like \astropy, what happens
depends mostly on what is highest on the priority list of those that
have time to contribute).

\paragraph{Uncertainties and Distributions.} Early in our carreers, we
were likely all taught that a measurement without a unit or an
uncertainty is essentially meaningless.  \astropyunits already
provides the ability to associate numbers with units, and propagate
these correctly, with its \astropyQuantity\ class, but so far the
ability to associate and propagate uncertainties is relatively
limited: in \astropysubpkg{nddata}, there are options to associate
errors with data arrays, but these are treated as independent, i.e.,
covariances are not tracked.

In general, proper error propagation is difficult, and often it is
best to use Monte Carle methods.  With the new
\astropysubpkg{uncertainties} sub-package, a first step is made to
make this possible.  It allows one to generate, for each variable,
randomly drawn samples in a \astropyDistribution, and then propagate
these by passing them through the normal analysis, producing a
\astropyDistribution\ of final results that can be inspected.

This \astropyDistribution\ can be any type of array, including of
course a \astropyQuantity, and the goal is to ensure it can be
seamlessly used to instantiate also other astropy classes, such as
\astropySkyCoord\ and \astropyTime.

Apart from Monte-Carlo error propagation, the plan for
\astropysubpkg{uncertainties} is to also support the case where values
have associated uncertainties, and these are propagated assuming that
they are normally distributed, but tracking covariances.  One
implementation problem to be solved is when to {\em stop} tracking
covariances.  For instance, a simple operation such as subtracting the
mean from $N$ data points implies that all data points are now
covariant with each other, i.e., one has to carry $N\times N$
covariances.  For a large image, that becomes not just pointless but
also prohibitive in terms of memory and CPU usage.

\paragraph{Masked quantities.}  Data can be known to be bad, in which
case it is useful to mask them.  Support for masked data involves a
number of choices, such as whether one indicates data values are bad
by a separate flag or by replacing them by a special value, such as
``Not a Number'' (NaN).  In \astropypkg, both approaches have been
used: a flag for masked columns in \astropyTable, mask flags and
bitmaps for N-dimensional data in \astropysubpkg{nddata}, and
replacing elements with NaN in \astropyTime.  A particular problem was
that no masked quantities were available, making it more difficult to
migrate to \astropyQTable, in which \astropyQuantity\ is used for all
columns with units, thus propagating those correctly (instead of
treating the unit merely as metadata).

Unfortunately, it turned out that the \texttt{MaskedArray} class from
\package{numpy} could not be used, as its design made it a wrapper
around any subclass, rather than allow another class to use it as an
alternative to a regular data array.  For quantities, this does not
work, as one can see from the fact that a masked value does not mean
the unit is not known.  Given this, a new \astropyMasked\ class was
designed, based on a framework very similar to that of
\astropyDistribution, making it easy to create masked instances of
other classes.  In particular, this can be used to create masked
quantities for \astropyQTable, including full support for reading and
writing to the different file types.

The masked quantities work mostly without any changes in higher-level
objects such as \astropySkyCoord, but work is still underway to expose
the mask in those classes, as well as to use masked arrays in
\astropyTime\ instead of replacing data with NaN.

\paragraph{Time series.}  From sampling a continuous variable at fixed
times to counting events binned into time windows, many different
areas of astrophysics require the manipulation of time series.  These
are often logically tables, and the new \astropysubpkg{timeseries}
sub-package extends the \astropyQTable\ class to tables of data as a
function of time, where the data can either represent samples or
averages over particular time bins.  The new classes offer a number of
special methods to manipulate time series (folding, resampling, etc.)
and to read different data formats (such as {\em Kepler} lightcurves,
etc.).

Also part of \astropysubpkg{timeseries} are common analysis routines,
including Lomb-Scargle and box-least-squares periodogrammes.

\paragraph{Spectral Coordinates.} Measurements are often taken at
specific ``spectral coordinates'', be they frequencies, wavelengths,
or photon energies.  The new \astropySpectralCoord\ class allows to
transform between these as well as to equivalent Doppler velocities.
It builds on the \astropyQuantity, which can also represent these
kinds of physical quantities, and allows conversion via dedicated
equivalencies, but aims to make this more straightforward.
Furthermore, it can be made aware of the observer and target reference
frames, allowing for instance transformation from telescope-centric (or
topocentric) frames to Barycentric or Local Standard of Rest
(LSRK and LSRD) velocity frames.

\paragraph{A High-Level Interface to World Coordinate Systems.}
Astronomical data are often provided with information about the
``real-world'' coordinates that correspond to pixel coordinates. This
mapping is the essence of the ``World Coordinate System (WCS)''
concept.  From its inception, the \astropysubpkg{wcs} sub-package
allowed access to WCS information provided in, e.g., FITS files, but
it became clear that other WCS standards and representations had to be
supported for new missions and observatories (e.g., the James Webb
Space Telescope and the Rubin Observatory).  To harmonize these, a new
high-level interface was created, based on a formal design first
proposed via an \astropy\ Proposal for Enhancement (\astropyAPE{14}).
The hope is that by having a formal design, it will be straightforward
for other packages implementing WCS objects to either modify their
classes to conform to the new interface or build thin wrappers that
conform to it.  The implementation in \astropysubpkg{wcs} interacts
well with other \astropypkg\ objects such as \astropySkyCoord\ and
\astropyTime.

\subsubsection*{Interoperability} \label{sec:core-features-interoperability}

  % New features capture a lot of attention. While exciting, the arguably more
  % significant improvements are those of interoperability -- ensuring existing
  % features connect smoothly together.

  The modular nature of \astropypkg means that not all sub-packages deeply
  integrate with one another -- and similarly between separate \python
  libraries. Because \astropy is situated at the nexus between scientific
  computing into astronomy, interoperability has been and continues to be an
  important focus for \astropy, both within the various \astropypkg sub-packages
  and with other important \python libraries. While increasing inter and
  intra-package integration will be a perpetual goal -- with feature additions
  in one place requiring updates in another -- \astropy has made impactful
  progress with units, tables, and I/O.
  % Better list? Also better transition
  If done right, all a user notices is that code using many different python
  features and libraries ``just works''. Here we will draw notice to these
  efforts to make your code ``just work''.

  \paragraph{\package{numpy} on Units}

    \astropyQuantity\ is a backbone of \astropy, leveraging the power of
    \package{numpy} and adding units. Previously, many \package{numpy} functions
    would strip a \astropyQuantity\ of its units (or fail outright), limiting
    \astropyQuantity's usefulness. Now, advances in \package{numpy} function
    overloading (see for example
    \href{https://numpy.org/neps/nep-0018-array-function-protocol.html}{NEP 18})
    mean \astropyQuantity\ works with almost all \package{numpy} (version 1.17+)
    functions. These advances have similarly benefitted many
    \package{numpy}-dependent packages, like \package{scipy} and their sparse
    matrices. And again \astropypkg benefits since more \package{scipy}
    functions now work with \astropyQuantity.

    In the remaining gaps, the \package{numpy} and \astropyQuantity\
    interoperability efforts are ongoing. For some functions, such as in
    \href{https://numpy.org/doc/stable/user/basics.rec.html#module-numpy.lib.recfunctions}{\package{numpy}'s
    module} for manipulating structured arrays, compatibility only requires
    exending the existing \package{numpy}-\astropypkg bridge framework.
    Community interest, in the form of a Feature Request (or better yet Pull
    Request) would be sufficient to see this realized. For other functions,
    discussed further in the
    \href{https://docs.astropy.org/en/stable/known_issues.html#known-deficiencies}{\astropy
    documentation}, the \package{numpy} framework does not yet allow for full
    interoperability with \astropy. A goal of \astropy, and the scientific
    \python community is to enhance and implement the frameworks, allowing
    \astropyQuantity\ to propagate units seamlessly across the whole ecosystem of
    \package{numpy}-like projects.

  \paragraph{Astropy in Time and on Units}

    % Bad sentence ↓. Rewrite.
    Not only did \astropyunits gain powerful integrations with \package{numpy},
    there were many intra-package improvements.

    \astropymodeling can add units to otherwise unit-less models.

    A new module –
    \astropycosmologyunits
    – has been added to the cosmology sub-package for defining and collecting
    cosmological units and equivalencies. The unit \texttt{littleh} and
    equivalency
    \astropyapidoc{cosmology.units.with_H0}{with\_H0}
    are deprecated from the main \astropyunits sub-package and moved to
    \astropycosmologyunits. A new unit, \texttt{redshift}, is added for tracking
    factors of cosmological redshift.
    % As this is a pseudo-unit an equivalency
    % \astropyapidoc{cosmology.units.dimensionless_redshift}{dimensionless\_redshift}
    % is added (and enabled by default) to allow for redshift - dimensionless
    % conversions.
    To convert between redshift and other cosmological distance
    measures, e.g. CMB temperature or comoving distance, the equivalency
    \astropyapidoc{cosmology.units.with_redshift}{with\_redshift}
    is also added. This equivalency is actually a composite of other
    equivalencies,
    % : \astropyapidoc{cosmology.units.redshift_distance}{redshift\_distance},
    % \astropyapidoc{cosmology.units.redshift_hubble}{redshift\_hubble}, and
    % \astropyapidoc{cosmology.units.redshift_temperature}{redshift\_temperature},
    which may be used separately.

    % IN PROGRESS (ADAPTING FROM THE WHATSNEW).
    The FITS Time standard paper defines the formats and keywords used to
    represent timing information in FITS files. The Astropy FITS package now
    provides support for reading and writing native Time table columns using
    this format. This is described within the FITS unified I/O interface and
    examples of usage can be found in the Astropy native objects (mixin columns)
    section.

    Implementation of this feature was made possible through the support of the
    Google Summer of Code 2017 program. To learn more about this project please
    see the final report, A mixin protocol for seamless interoperability.

  \paragraph{Table Mixin Columns}

    Numerous \python libraries have custom table implementations
    \begin{itemize}
      \item (Support use of SkyCoord as Column happened before) and in table vstack, dstack, and insert\_row
      \item Converting SkyCoord to QTable (1 sentence addition to above)
      \item Added support for dask arrays in tables
      \item Added support for registering array-like objects as mixin columns
    \end{itemize}

  \paragraph{Unified I/O architecture}

    \astropyiosubpkg{registry} is a powerful way to define input and output
    (I/O) functions for \astropypkg objects, such as reading or writing to a
    file. \astropy uses this internally for the I/O methods in \astropyTable and
    \astropyCosmology (I/O added v5.0+), and users can register custom I/O to
    extend the options on these classes.

    As of Astropy 5.0 the I/O registry submodule has switched to a class-based
    architecture and been refactored to allow read-only, write-only, and
    read-write registries. Additionally, the registry submodule no longer
    assumes an object has \astropy-style \texttt{read} and \texttt{write}
    methods when trying to read or write that object. These two improvements
    enable a number of new use cases for \astropyiosubpkg{registry}, some
    highlighted here, but others we hope to see have community adoption.

    First, the class-based architecture allows for the creation of custom
    registries. An application of this is in \astropycosmology, which
    has two different \textit{kinds} of registries for \astropyCosmology: one
    for reading and writing files, and another for converting between python
    objects.

    Second, \textit{any} kind of object is now compatible with
    \astropyiosubpkg{registry}. Using \astropyiosubpkg{registry} for custom
    classes means a user does not have to create a whole new I/O system and can
    offer a unified and familiar API.

  \paragraph{Persistent Storage and I/O}

    % TODO Better transition sentence
    The imrovements to \astropy's I/O architecture are mirrored in the increased support for I/O options.

    \begin{itemize}
      \item ASDF : JSWST  and models can be stored with ASDF.
      \item ECSV : Support for multidimensional and object columns in ECSV
      \item FITS : Support for writing Dask arrays to FITS files, and Append table to existing FITS file
      \item Support for reading, writing, and converting Cosmology
      \item MRT Support for reading and writing tables to MRT format : GSOC
      \item Support for reading and writing tables to Parquet format
      \item Support for reading and writing tables to QDP format
      \item ASDF read write
    \end{itemize}

\subsection*{Precision, Accuracy, \& Reproducibility} \label{sec:core-features-precision_accuracy_repreducibility}

  An important focus of \astropy has been to increase the precision and accuracy
  of its functionality, while, where appropriate, making sure results obtained
  using older versions of \astropypkg are reproducible.
  % Something about ScienceState
  Notable improvements have been made to \astropytime, \astropyconstants,
  \astropycoordinates, and \astropycosmology.

  \paragraph{Time}

    In astronomy, time accuracy down to the second is frequently important.
    Whether for planning observations or crunching data, missing seconds can
    meaningfully impact results. Before, when using \astropyTime in \astropytime
    leap seconds had to be manually applied; now leap seconds are applied
    automatically. Moreover, \astropypkg updates internal time data-files to
    ensure that the correct leap second adjustment is always used. For
    reproducibility, e.g. with an old code, \astropyLeapSeconds may also be
    manually applied.

  \paragraph{Constants}

    Measurements of physical constants (and the units defined from them) improve
    over time. Periodically the standardized systems of units and constants are
    updated to reflect these improvements, for instance in 2019 the SI system was
    redefined \citep{NIST2019}, with an accompanying update to the physical
    constants in
    \href{https://codata.org/initiatives/data-science-and-stewardship/fundamental-physical-constants/}{\texttt{SI/CODATA
    2018}}. \astropypkg contains two modules -- \astropyconstants and
    \astropyunits -- for working with constants and units, respectively.
    \astropyconstants now defaults to use the \texttt{SI/CODATA 2018} values, with
    the relation between units in \astropyunits based upon these constants. Most
    of \astropypkg and affiliate packages build upon \astropyunits, so this update
    has far-reaching effects across the \astropy ecosystem and astronomy
    community.

    For reproducibility, \astropypkg allows the constants (and therefore units)
    definitions to be rolled back to prior values, for instance the
    \texttt{SI/CODATA 2014} definitions.
    % CITATION
    The constants and units definitions are set by configuration file as well as
    \astropyScienceState (an runtime configuration controller). For work where
    updates to the values of the fundamental constants significantly alters
    results we recommend including with the work an astropy configuration file,
    which specifies the set of constants used.

  \paragraph{Cosmology}
    % TODO! At least one good sentence about "what is the benefit of astropy.cosmology"

    \astropycosmology contains classes for representing cosmological models.
    % currently all homogeneous and isotropic variations of the
    % Friedmann-Lemaitre-Robertson-Walker metric.
    Bundled with the classes are commonly used \astropyCosmology realizations, for
    example best-fit measurements from
    \href{https://lambda.gsfc.nasa.gov/product/map/current/}{WMAP}
    \citep{WMAP2003} and \href{https://www.nasa.gov/mission_pages/planck}{Planck}
    \citep{PlanckMission:2006}.

    When these important missions publish new measurements, \astropycosmology is
    updated to include these results as realizations of the appropriate class
    (generally \astropyFlatLambdaCDM). The models used by WMAP and Planck do not
    always exactly correspond to \astropycosmology classes, for instance the
    Planck 2018 results \citep{Planck2018VI} include massive neutrinos in
    $\Omega_{matter,0}$, while in \astropyFlatLambdaCDM this mass contribution is
    stored in a separate parameter. Consequently, while some parameter values
    appear different from the source paper, the \astropyCosmology realizations
    correctly reproduce the cosmological models.

    Since \citealt{astropy:2018} the following cosmology realizations have been
    added:
    \href{http://docs.astropy.org/en/stable/api/astropy.cosmology.WMAP1.html}{\texttt{WMAP1}}
    from WMAP First Year \citep[Table 7: WMAP + CBI + ACBAR + 2dFGRS +
    Lya]{WMAP1Year:2003};
    \href{http://docs.astropy.org/en/stable/api/astropy.cosmology.WMAP3.html}{\texttt{WMAP3}}
    from WMAP Three Year \citep[Table 6: WMAP + SNGold]{WMAP3Year:2007};
    \href{http://docs.astropy.org/en/stable/api/astropy.cosmology.Planck15.html}{\texttt{Planck15}}
    from Planck 2015 \citep[Table 4: TT, TE, EE + lowP + lensing +
    ext]{Planck2015XIII:2016};
    \href{http://docs.astropy.org/en/stable/api/astropy.cosmology.Planck18.html}{\texttt{Planck18}}
    from the Planck 2018 best-fit cosmological parameters \citep[Table 2: TT, TE,
    EE + lowE + lensing + BAO]{Planck2018VI:2020}.

    Astropy provides a
    \href{http://docs.astropy.org/en/stable/api/astropy.cosmology.default_cosmology.html}{configurable
    default cosmology}. All calculations done in a cosmological context will use
    this default cosmology. The default \astropyCosmology has been updated to the
    Planck 2018 parameters. For reproducibility, the default cosmology may be set
    to old defaults, such as the WMAP Three Year values. The default cosmology is
    dynamically configurable using \astropyScienceState, meaning a set of
    calculations may be run using different assumed cosmologies, and results
    compared between the two.
    % Repeat the word default less.

  \paragraph{Coordinates}

    Being able to describe the position of objects, i.e. their coordinates, is
    fundamental across many knowledge domains. \astropypkg provides
    \astropycoordinates for working with low-level positional and velocity data
    all the way to high-level coordinate objects with reference frames,
    atmospheric information, and more. A central feature of \astropycoordinates is
    the ability to transform data between reference frames, e.g. \astropyICRS
    \citep{ICRS:1997} to \astropyGalacticLSR \citep{GalacticLSR:2010}. However,
    there have been some historical limitations for spatially proximate
    transformations, impacting the usefulness of \astropycoordinates for
    ground-based telescopes and intra-solar astronomy. Transformations in the
    \astropyAltAz frame were reasonably precise for very distant objects, but
    wrong by up around several arcseconds for e.g. the location of the moon. Now
    these transformations are much more precise, down to the milli-arcsecond
    level. Similar precision improvements were made to the
    \astropyapi{coordinates.builtin_frames.HADec}{Hour Angle-Declination} frame
    transformations. Additionally,
    \astropyapi{coordinates.builtin_frames.BaseEclipticFrame}{Ecliptic frames}
    and associated transformations have been updated to correctly reflect the
    “true” (includes precession but not nutation) and “mean” (precession and
    nutation) terminology.

    For Galactic astronomers, the \astropyGalactocentric
    \astropyapi{coordinates.galactocentric_frame_defaults}{frame defaults} have
    been updated to include more recent measurements (tabulated in the
    \href{https://docs.astropy.org/en/stable/api/astropy.coordinates.galactocentric_frame_defaults.html#astropy.coordinates.galactocentric_frame_defaults.references}{\texttt{frame\_defaults.references}}
    attribute). For reproducibility, old definitions are still available, and will
    continue to be so as \astropypkg is updated with forthcoming measurements of
    the Galactic center, solar motion, etc.
    % may be used by setting the default value to ``pre-v4.0'' in the
    % \href{https://docs.astropy.org/en/stable/api/astropy.coordinates.galactocentric_frame_defaults.html}{\texttt{galactocentric\_frame\_defaults}}
    % \astropyScienceState.

\subsection*{Performance} \label{sec:core-features-performance}

As features are added and corner cases dealt with, performance often
suffers.  Hence, pull requests that improve performance are welcomed
generally, and were even the main goal for one release cycle.  For
most packages, this meant improvements in the \python\ code, but for
some the most time-critical pieces were rewritten in~C (e.g., for
convolution of images, sigma-clipping, and converting time strings to
binary).  Furthermore, a particular effort was made to ensure
\astropypkg\ is thread-safe, so that it can be used on supercomputer
clusters.

% \subsection*{Future} \label{sec:core-features-future}
% MHvK: Merged with the new feature section

@@

>>>>

% Links to documentation: subpackages
\newcommand{\astropysubpkg}[1]{\href{http://docs.astropy.org/en/stable/#1/index.html}{\texttt{astropy.#1}}\xspace}
\newcommand{\astropyiosubpkg}[1]{\href{http://docs.astropy.org/en/stable/io/#1/index.html}{\texttt{astropy.io.#1}}\xspace}
\newcommand{\astropywcsaxes}{\href{http://docs.astropy.org/en/stable/visualization/wcsaxes/index.html}{\texttt{astropy.visualization.wcsaxes}}\xspace}

% Links to documentation: classes
\newcommand{\astropyskycoord}{\href{http://docs.astropy.org/en/stable/api/astropy.coordinates.SkyCoord.html}{\texttt{SkyCoord}}\xspace}
\newcommand{\astropyQuantity}{\href{http://docs.astropy.org/en/stable/api/astropy.units.Quantity.html}{\texttt{Quantity}}\xspace}
\newcommand{\astropyTime}{\href{http://docs.astropy.org/en/stable/api/astropy.time.Time.html}{\texttt{Time}}\xspace}
\newcommand{\astropyTable}{\href{http://docs.astropy.org/en/stable/api/astropy.table.Table.html}{\texttt{Table}}\xspace}

@@

Summarize major pieces of development since the last paper (since v2.0). A few
ideas (taken from What's New pages) below, but feel free to expand on or remove
from this list:
\begin{itemize}
    \item Support for representing and transforming velocity data in coordinates, epoch propagation (v3.0)
    \item Improved support for astronomical time series: TimeSeries object (v3.2), Box Least Squares periodogram (v3.1)
    \item Overall improved support of Quantity throughout numpy (v4.0) and scipy
    \item Native support for Time, Quantity, and SkyCoord objects in Astropy tables
    \item APE 14 WCS
    \item A new SpectralCoord class for representing and transforming spectral quantities
    \item Support for dask/large arrays in several places in the core package (4.1, 5.0, and maybe other versions)
\end{itemize}

@@

All coordinated and affiliated packages are listed on the astropy
website\footnote{\url{https://www.astropy.org/affiliated}}. Since 
\cite{astropy:2018}, there has been an expansion of affiliated
packages for gravitational astrophysics including:
\texttt{PyCBC} for exploring gravitational wave signals, lenstronomy for
modeling strong gravitational lenses, \texttt{ligo.skymap} for visualizing
gravitational wave probability maps, and \texttt{EinsteinPy} for general
relativity and gravitational astronomy. There have been several packages
added to the ecosystem related to HEALPix: \texttt{astropy-healpix}
for a BSD licensed HEALPix implementation, and \texttt{mocpy} for Multi-Order 
Coverage maps. \texttt{astroalign} has been introduced for astrometric registration,
and \texttt{python-cpl} has been added for ESO pipelines and VLT data products. 
For ground-based astronomy, \texttt{baseband} has added IO capabilities for 
VLBI, and \texttt{SpectraPy} brings slit spectroscopy to the astropy ecosystem. 
We have \texttt{agnpy} for AGN astrophysics, and \texttt{statmorph} for fitting galactic 
morphological diagnostics. \texttt{saba} gives an interface to sherpa's fitting 
routines. For modeling interstellar dust extinction we have 
\texttt{dust_extinction}, and for extracting features from time series we 
have \texttt{feets}. We also have \texttt{corral} for managing data intensive parallel 
pipelines. \texttt{BayesicFitting} provides an interface for generic Bayesian 
inference, and \texttt{sbpy} does calculations for asteroid and cometary 
astrophysics. Finally, \texttt{synphot} provides an interface for synthetic photometry.

Several of the coordinated and affiliated packages described in [reference to
paper 2] have had substantial improvements. \texttt{astroquery}
\cite{astroquery} has added
access to roughly a dozen new missions and data services, including JWST, and
the project has switched to a continuous release model. Every time a change
is committed to the main development branch it is published on the Python
Package Index (PyPI, ref) and available for installation. Formal releases are
still done a couple of times per year. \texttt{photutils} \cite
{photutils} released its first stable version, indicating that the API will
change less frequently, and there have been several significant performance
improvements.  \texttt{ccdproc} \cite{ccdproc} also released a new major
version, bringing better performance to some image combination operations.
The \texttt{regions} package \cite{pyregions}, for manipulating ds9-style
region definitions \cite{ds9}, added new ways to manipulate regions and
introduced new region types. \texttt{reproject}’s \cite{reproject} major new
feature is a function to align and co-add images to create a mosaic; better
support for parallelization was also added. \texttt{specutils}, the package
that defines containers for 1D and 2D spectra \cite{specutils}, also had its
first stable release and the addition of classes to read JWST data.
\texttt{stingray} has also had a major performance overhaul.
